---
title: "class11"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##PDB statistics

Q1: What proportion of PDB entries does X-ray crystallography account for? What proportion of
structures are protein?
89.51%
```{r}
#import our PDB statistics CSV file and calculate percent structures by experimental method
pdbstats <- read.csv("Data Export Summary.csv", row.names = 1)
#Calculate and assign numbers with a category
percent <- (pdbstats$Total / sum(pdbstats$Total))*100
names(percent) <- row.names(pdbstats)
percent

```



Q2: Type HIV in the search box on the home page and determine how many HIV-1 protease
structures are in the current PDB?
118


Q3: Water molecules normally have 3 atoms. Why do we see just one atom per water molecule
in this structure?
We only see oxygen atom. Hydrogen atoms are too small to be visualized.


Q4: There is a conserved water molecule in the binding site. Can you identify this water
molecule? What residue number does this water molecule have (see note below)?


```{r}
library(bio3d)
```

Read in out HIV protease

```{r}
pdb <- read.pdb("1hsg")
pdb
```

Q6. How many amino acid residues are there in this pdb object and what are the two nonprotein
residues?

198 aa residue.
Non-protein/nucleic resid values: [ HOH (127), MK1 (1) ]


Q7. What type of R object is pdb$
```{r}
attributes(pdb)
str(pdb$atom)
```

#ATOM selection in PDB

```{r}
inds.ligands <- atom.select(pdb,"ligand")
inds.protein <- atom.select(pdb,"protein")

```
```{r}
# Print a subset of $atom data for ligands
pdb$atom[inds.ligands$atom,]
```


```{r}
# Print a subset of $atom data for protein
pdb$atom[inds.protein$atom,]

```

```{r}
head(pdb$atom[inds.protein$atom,])
```

#select protein only atoms and write
```{r}
pdb.ligand<-trim.pdb(pdb,inds=inds.ligands)
pdb.ligand
write.pdb(pdb.ligand,file="1hsg_ligand.pdb")

pdb.protein <-trim.pdb(pdb,inds=inds.protein)
pdb.protein
write.pdb(pdb.protein,file="1hsg_protein.pdb")
```

Q8. Use the Bio3D write.pdb() function to write out a new C-alpha atom only PDB file for
viewing in VMD. 
```{r}

# Select C-alphas 
ca.inds <- atom.select(pdb, "calpha")
pdb.ca <-trim.pdb(pdb,inds=ca.inds)
pdb.ca
write.pdb(pdb.ca,file="1hsg_ca.pdb")

```



#muscle

test muscle by aligning multiple pdb
```{r}
# Download some example PDB files
ids <- c("1TND_B","1AGR_A","1TAG_A","1GG2_A","1KJY_A","4G5Q_A")
files <- get.pdb(ids, split = TRUE)
# Extract and align the chains we are interested in
pdbs <- pdbaln(files, fit = TRUE)
# Print to screen a summary of the 'pdbs' object
pdbs
```

#BLAST SEARCH PDB


```{r}
aa <- get.seq("1ake_A")
b <- blast.pdb(aa)

```
```{r}
# Plot a summary of search results
hits <- plot(b)
```


```{r}
hits
head(hits$pdb.id)
```

```{r}
# Fetch PDBs
files <- get.pdb(hits$pdb.id, path = "pdbs", split = TRUE, gzip =
TRUE)
# Align structures
pdbs <- pdbaln(files)
# Vector containing PDB codes
ids <- basename.pdb(pdbs$id)
# Draw schematic alignment
plot(pdbs, labels=ids)
```


sequence conservation analysis
```{r}
# Calculate sequence conservation
cons <- conserv(pdbs, method="entropy22")
# SSE annotations
sse <- pdbs2sse(pdbs, ind=1, rm.gaps=FALSE)
# Plot conservation per residue
plotb3(cons, sse=sse, ylab="Sequence entropy")
```


principal component analysis
```{r}
# find invariant core
core <- core.find(pdbs)
# superimpose all structures to core
pdbs$xyz = pdbfit(pdbs, core)
# Perform PCA
pc.xray <- pca(pdbs)
# Calculate RMSD
rd <- rmsd(pdbs)
# Structure-based clustering
hc.rd <- hclust(dist(rd))
grps.rd <- cutree(hc.rd, k=3)
plot(pc.xray, 1:2, col="grey50", bg=grps.rd, pch=21, cex=1)
```

