---
title: "bioinformaticsClass7"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Functions again

We can source any file of R code with the 'source'() function

```{r}
source("http://tinyurl.com/rescale-R")
```

let's make sure things are here

```{r}
ls()
```

Check our rescale function is working

```{r}
rescale(1:10)
```
And the first error would be where you have things other than numeric values. 
```{r,eval=FALSE}
rescale(c(1:10,"string"))
```

We can add a if loop for a warning message
This function is stored in rescale2
Lets see if rescale2 is better

```{r,eval=FALSE}
rescale2(c(1:10,"string"))
```

How to compile when error message is here?

1) Comment the error chunks
2) Add ",eval=FALSE"" after the beginning r code chunk after r in the parenthesis

##Function for finding missing values in 2 datasets
Write a "both_na()" function to do this
```{r}
x <- c( 1, 2, NA, 3, NA)
y <- c(NA, 3, NA, 3, 4)
 
is.na(x)

```

```{r}

which (is.na(x))
```
You can flip your logical choice by "!"

```{r}
which (!is.na(x))
#your "!" flipped your results
```


```{r}
z<-10
if(z>5) {
  print("more")
} else {
  print("less")
}
```

```{r}
sum(is.na(x))
# This only tells you how many missing values are here
```

```{r}
is.na(x)
is.na(y)
```
Now we want to know when both are TRUE, and how many times that happen

```{r}
is.na(x) & is.na(y)
sum(is.na(x) & is.na(y))

```
Write it into a function
```{r}
both_na <- function(x,y) {
  sum(is.na(x) & is.na(y))
} 

```

test it
```{r}
both_na(x,y)
```

But you will get error when your input length doesn't match
```{r,eval=FALSE}
x <- c(NA, NA, NA)
y1 <- c( 1, NA, NA)
y2 <- c( 1, NA, NA, NA)

both_na(x,y1)
both_na(x,y2)
```

```{r,eval=FALSE}
both_na2(x,y2)

```

```{r}
x <- c( 1, 2, NA, 3, NA)
y <- c(NA, 3, NA, 3, 4)
ans <- both_na3(x,y)
```

```{r}
ans$which
```

##And a last example function

```{r}
x<- df1$IDs
y<- df2$IDs

x
y
# Using intersect we can find intersect components
intersect(x,y)

# Using value matching function returns positions of matching of x in y
match(x,y)

#Check if everything in x is in y
x %in% y
```
We can use the logical output of %in%
Combine the values with cbind.

```{r}
x[x %in% y]
y[y %in% x]


cbind(x[x %in% y],y[y %in% x])
```


```{r}
cbind(c("hello", "help"), c("please", "me"))
```
Notice that rbind binds them as rows
```{r}
rbind(c("hello", "help"), c("please", "me"))

```


Lets make them into a function

```{r}

intersectgenes <- function(x,y) {
  cbind(x[x %in% y], y[y %in% x])
}

intersectgenes(x,y)
```

Our x and y are in a data frame
```{r}
#This is my code
intersectgenes2 <- function(df1,df2) {
  cbind(df1[ df1$IDs %in% df2$IDs, ], df2[df2$IDs %in% df1$IDs, "exp"])
}

intersectgenes2(df1,df2)
```

```{r}
#This is professor's code
gene_intersect2 <- function(df1, df2) {
 cbind( df1[ df1$IDs %in% df2$IDs, ],
 df2[ df2$IDs %in% df1$IDs, "exp"] )
}
```

But now the column name looks complex

```{r}
intersectgenes3 <- function(df1,df2) {
  gene.colname="IDs"
  cbind( df1[ df1[,gene.colname] %in% df2[,gene.colname], ],
 exp2=df2[ df2[,gene.colname] %in% df1[,gene.colname], "exp"] )
}

intersectgenes3(df1,df2)
```
Change a bit about the words

```{r}
intersectgenes4 <- function(df1, df2, gene.colname="IDs") {
 df1.name <- df1[,gene.colname]
 df2.name <- df2[,gene.colname]
 df1.inds <- df1.name %in% df2.name
 df2.inds <- df2.name %in% df1.name
 cbind( df1[ df1.inds, ],
 exp2=df2[ df2.inds, "exp"] )
}

intersectgenes4(df1,df2)
```

If we want to use every column
Lets use merge function for this

```{r}
merge(df1,df2,by="IDs")

```

